How to write a PhD in Biological Sciences: a guide for the uninitiated

### 2021-07-20 1.03-lifescientific.Rmd
Add files via upload
# After copy edits

Throughout the book, I make reference to the ‘scientific project’ as a greater plan to which our studies in biological sciences is simply a part. I consider the scientific project as a path that humanity should continue to travel in order to reach a higher and more just way of living. Our societies have evolved from places where superstition and belief were the basis for societal rules that favoured the privileged classes. Science is a process in which we reach conclusions based on egalitarian reasoning and evidence, something that we can aspire for the larger community around us. Already, most societies are making the majority of their collective evidence based decisions through the guidance of science. Therefore, if we consider our knowledge based societies to be democratic, then all citizens of that society should understand the processes under which the foundation of that knowledge is acquired [@sagan2011demon]. Yet globally, public skepticism in science has been rising, and has reached as high as 35%. To consolidate this growing position of science in society, and the decline in public acceptance, scientists will need to become more open, communicative and seen to be conducting their own profession in a way that reflects these ideals. There is evidence that increasing communication from scientists to the public results in a decrease of skepticism ([see here](https://multimedia.3m.com/mws/media/1898512O/3m-sosi-2020-pandemic-pulse-global-report-pdf.pdf)).

Imagine a world in which all of the principle decisions are made as a result of advances in science, but vanishingly small numbers of people actually understand how science works. This is not a world that we would want, but as Carl Sagan [@sagan2011demon] pointed out, it’s one that could all too easily happen. Therefore, to prevent the workings of science falling into the hands of a minority, we have a duty to communicate our science to the majority.

Ultimately, what we want from the scientific project is to draw everyone in. That is not to say that there is no place in the world for other subject areas. One of the greatest misadventures in my own education was the indoctrination that I received that there were two streams in education, one for STEM and one for arts. I hope that you managed to escape this narrow minded and factually incorrect view, because as you will discover there are high levels of creativity demanded within the scientific framework, and those who have more talent in communicating their science will likely be more successful in their careers as a result. Moving forward with the scientific project does not mean that everyone must become a scientist, but instead that the process by which scientists establish knowledge is understood. Indeed, as we shall see (below) sceptical thinking, as exemplified in science, is indeed a healthy way for everyone in society to avoid time wasting and potentially being deceived or conned. It is also vital that the scientific process be transparent in a way that results can be understood and therefore respected.

@@

## Example: Invasive species 
Alien species are those that have been moved by human agents across biogeographic barriers into parts of the world where they did not evolve. Once these aliens arrive in their new environment, some breed and become established, and after a lag a fraction of these spread and become invasive, often with substantial impacts on the environment and the socio-economic efforts of people in those invaded areas. Invasive species disproportionately affect poorer people in our society, often completely disassociated with the pathways responsible for the introduction and spread. Removing invasive species requires a considerable amount of effort, and unless this is supplied by tax-payers, via governmental policy, falls on those directly impacted. Invasion science, the expanded transdisciplinary area stemming from invasion biology, is required to engage with all stakeholders in invasions, and most importantly to prevent new invasions. Because introductions are made by humans there is a critical need to communicate with those stakeholders along introduction pathways.

For existing invasions, social scientists and economists are needed to determine their impacts on people in invaded areas and determine how to best alleviate the invasion or mitigate its effects without it moving further and impacting more communities. Biological scientists need to assess the impact on the environment, and investigate ways in which the species can be removed, their spread curtailed, or their impacts neutralised. Importantly, the results need to be discussed with all stakeholders to achieve a desirable outcome for those impacted most. Invasions at larger scales will normally require governmental resources to ensure that communities can continue to function, especially to best deliver their services to the society. Governments require evidence based decisions that are reached from well researched studies conducted in the field. Effective measures may necessitate large sums of money, and governments require that this process be transparent so that the taxpayers’ money they are investing is spent efficiently.

@@

## Should society lead the sciences or sciences lead society? 
Of course, the answer is simply that there must be feedback between science and society. As science advances, society learns of these advances and views are advanced. To become incorporated as a societal norm takes time. Over time, views from science that appear radical to society become mainstream and accepted. Think of the views on the earth being round, or more recently the general acceptance that the universe started with the ‘big bang’. The counterculture, also known as post truth, faction in society that espouses views accepted by the mainstream and proved by science beyond reasonable doubt is not a novel phenomenon. Indeed, in the scientific project, we need to accept that these views will persist, but empower the general populace to understand why and how they can be shown to be false. Hence, your view that the earth is not flat is bolstered if you can provide three ways to demonstrate this as fact (pick the ones you like from 7 listed [here](http://crosstalk.cell.com/blog/seven-ways-to-prove-earth-is-round) or [here](https://www.unlv.edu/news/release/round-earth-clues-how-science-proves-our-home-globe)). Otherwise, your assertion that it is not flat will appear as easily denied as a ‘flat-earther’s’ assertion that it is. The power of the scientific argument in society is that it is not simply contradiction or gainsay, but that there is evidence.

The scientific project is currently not the open virtuous eutopia that we would like. Indeed, as scientists are humans and part of the current society, they are prone to the same trappings of wielding power and influence in order to get ahead. As you get toward the end of this book, you will see me calling out increasing parts of the scientific project that are currently riddled with bias, thereby weakening the entire process and disenfranchising the most vulnerable. Reforming science is no easy task. Reforms are needed throughout the structures that fund, employ and disseminate scientific research. Sweeping this all away will take a concerted effort, but might well be easier in science than in other sectors of society. Nearly all scientists recognise that science needs to be objective and impartial in order to succeed. Their objectivity makes them an ideal group to kickstart a fairer, more transparent and open system. As you will see, there has already been a move to make science more open, and by embracing this transparency the next generation of scientists will reap the benefits, and be on a sounder footing to improve a better scientific society.

@@

In his book, Demon-Haunted World, Carl Sagan [-@sagan2011demon] provides his take on how science provides scientists with what he terms ‘Baloney Detection’, commonly referred to today as ‘bullshit’. These are well worth recording here as they provide a guiding light for all scientists whether they be novices, emerging or established. They are also easily taught, and the foundation of what we should spread in the Scientific Project.

As humans, we are all susceptible to bullshit, hoaxes, charlatans and contrived deceptions as they are often more appealing than the truth. However, science is about truth seeking, and by applying a set of simple rules, Sagan argues, we can avoid being misled or distracted by ideas that may waste our time and energy by using ‘skeptical thinking’. These rules are useful in your scientific life when considering hypotheses or when building arguments to explain your results. However, as Sagan argues, these simple rules are also useful in your day to day life, and as a simplified example of the scientific method, we can pass them onto all our fellow citizens.

1. You should always seek independent confirmation of results. 
1. Seek out different opinions on the results and consider all possibilities
1. Don’t believe facts simply because they are spoken by an authority on the subject. We know that humans are prone to [confirmation bias](type1error.html), and often exhibit motivated reasoning, especially if they have some vested interest. Even the most experienced have made mistakes in the past and will again in the future. My preferred way of thinking of this rule is to acknowledge that experts are made of ex- (a has been) and -spurt (a drip under pressure).
1. [Use a hypothetical deductive framework](howhypothesis.html). Instead of accepting the first (hypothesis) explanation, think of all the possible reasons why a result might have occurred and then consider how you could (or perhaps you already have in your experiment) disproved each one. You may find that an untested hypothesis to explain the results represents a major caveat that could be tested, or can’t yet be tested. Either way, you need to remain open to the possibility that this explains your results. This leads neatly onto:
1. Don’t get too involved with keeping your chosen hypothesis [opening you to potential cofirmation bias](type1error.html). Remain critical to the possibility that it might be wrong, and provide ideas on how to substantiate or disprove competing hypotheses. If you don’t, others will. 
1. If at all possible use a quantitative method to test your hypothesis. You should already be well on the way to this one inside your PhD, but it is worth bearing in mind especially when it comes to considering testing other competing hypotheses. How could they be quantitatively tested?
1. Arguments ([as we see later](Argument.html)) consist of a chain of statements that follow one from the other. This chain is only as strong as its weakest link, and the strength of each link in the chain (your argumentative statements) must be critically considered by you. 
1. [Occam’s Razor](Argument.html) dictates that given two hypotheses that equally explain the results, the simpler one is likely more plausible. Or put another way, the more links you have in your chain, the more likely that one of them is not as strong as you would like it to be.
1. Your [hypothesis](howhypothesis.html) (or argument) must always be falsifiable. That is, that your test should be able to find that it is incorrect. If your test cannot show that your hypothesis is incorrect, then it isn’t worth much.

Sagan provides plenty of real world examples of each of these arguments, and I’d suggest that you consult his book [@sagan2011demon] if you are interested in following these up. Similar approaches, but written in different styles, are provided by Richard Dawkins in his [-@dawkins2004devil] book: A Devil’s Chaplain, and Julia Galef in her [-@galef2021scout] book: The Scout Mindset.

@@

I would encourage you not to stop your objective and evidence based thinking when you leave the office. By applying the rules of science to other sectors of society, you will find that there are many societal norms that could benefit from similar reform. Certainly, a lot of the technology developed in science can be used to help society reform, but scientific thinking will help everyone. As well as bringing the world to scientific thinking, we can take scientific thinking to the world.

## Citizen science
For a long time I struggled with the term 'Citizen science' as I reasoned that many such ventures did not involve citizens doing science. To me, most seemed focussed on outsourcing large tasks to willing public participants: i.e. having citizens do some grunt work. Collecting data, or doing a fraction of the analysis is not science. However, in light of the scientific project, I have changed my mind. All of these projects offer citizens a way to participate in scientific projects in a way that was previously unavailable to them. They promote direct communication between scientists and citizens. In this way, they serve to promote the understanding of science to the general public.

>>>>

Throughout the book, I make reference to the ‘scientific project’ as a greater plan to which our studies in biological sciences is simply a part. I consider the scientific project as a path that humanity should continue to travel in order to reach a higher and more just way of living. Our societies have evolved from places where superstition and belief were the basis for societal rules that favoured the privileged classes. Science is a process in which we reach conclusions based on egalitarian reasoning and evidence, something that we can aspire for the larger community around us. Already, most societies are making the majority of their collective evidence-based decisions through the guidance of science. Therefore, if we consider our knowledge-based societies to be democratic, then all citizens of that society should understand the processes under which the foundation of that knowledge is acquired [@sagan2011demon]. Yet globally, public skepticism in science has been rising, and has reached as high as 35%. To consolidate this growing position of science in society, and the decline in public acceptance, scientists will need to become more open, communicative and seen to be conducting their own profession in a way that reflects these ideals. There is evidence that increasing communication from scientists to the public results in a decrease of skepticism ([see here](https://multimedia.3m.com/mws/media/1898512O/3m-sosi-2020-pandemic-pulse-global-report-pdf.pdf)).

Imagine a world in which all of the principal decisions are made as a result of advances in science, but vanishingly small numbers of people actually understand how science works. This is not a world that we would want, but as Carl Sagan [@sagan2011demon] pointed out, it’s one that could all too easily happen. Therefore, to prevent the workings of science falling into the hands of a minority, we have a duty to communicate our science to the majority.

Ultimately, what we want from the scientific project is to draw everyone in. That is not to say that there is no place in the world for other subject areas. One of the greatest misadventures in my own education was the indoctrination that I received that there were two streams in education, one for STEM and one for arts. I hope that you managed to escape this narrow minded and factually incorrect view, because as you will discover there are high levels of creativity demanded within the scientific framework, and those who have more talent in communicating their science will likely be more successful in their careers as a result. Moving forward with the scientific project does not mean that everyone must become a scientist, but instead that the process by which scientists establish knowledge is understood. Indeed, as we shall see (below), sceptical thinking, as exemplified in science, is a healthy way for everyone in society to avoid time wasting and potentially being deceived or conned. It is also vital that the scientific process be transparent in a way that results can be understood and therefore respected.

@@

## Example: Invasive species 
Alien species are those that have been moved by human agents across biogeographic barriers into parts of the world where they did not evolve. Once these aliens arrive in their new environment, some reproduce and become established, and after a lag, a fraction of these spread, and become invasive, often with substantial impacts on the environment and the socio-economic efforts of people in those invaded areas. Invasive species disproportionately affect poorer people in our society, often completely disassociated with the pathways responsible for the introduction and spread. Removing invasive species requires a considerable amount of effort, and unless this is supplied by taxpayers, via governmental policy, falls on those directly impacted. Invasion science, the expanded transdisciplinary area stemming from invasion biology, is required to engage with all stakeholders in invasions, and most importantly to prevent new invasions. Because introductions are made by humans there is a critical need to communicate with those stakeholders along introduction pathways.

For existing invasions, social scientists and economists are needed to determine their impacts on people in invaded areas and determine how to best alleviate the invasion or mitigate its effects without it moving further and impacting more communities. Biological scientists need to assess the impact on the environment, and investigate ways in which the species can be removed, their spread curtailed, or their impacts neutralised. Importantly, the results need to be discussed with all stakeholders to achieve a desirable outcome for those impacted most. Invasions at larger scales will normally require governmental resources to ensure that communities can continue to function, especially to best deliver their services to the society. Governments require evidence-based decisions that are reached from well-researched studies conducted in the field. Effective measures may necessitate large sums of money, and governments require that this process be transparent so that the taxpayers’ money they are investing is spent efficiently.

@@

## Should society lead the sciences or sciences lead society? 
Of course, the answer is simply that there must be feedback between science and society. As science advances, society learns of these advances and views are advanced. To become incorporated as a societal norm takes time. Over time, views from science that appear radical to society become mainstream and accepted. Think of the views on the earth being round, or more recently, the general acceptance that the universe started with the ‘big bang’. The counterculture, also known as post-truth, faction in society that espouses views accepted by the mainstream and proved by science beyond reasonable doubt is not a novel phenomenon. Indeed, in the scientific project, we need to accept that these views will persist and empower the general populace to understand why and how they can be shown to be false. Hence, your view that the earth is not flat is bolstered if you can provide three ways to demonstrate this as fact (pick the ones you like from 7 listed [here](http://crosstalk.cell.com/blog/seven-ways-to-prove-earth-is-round) or [here](https://www.unlv.edu/news/release/round-earth-clues-how-science-proves-our-home-globe)). Otherwise, your assertion that it is not flat will appear as easily denied as a ‘flat-earther’s’ assertion that it is. Thus the power of the scientific argument in society is that it is not simply contradiction or gainsay but that there is evidence.

The scientific project is currently not the open virtuous eutopia that we would like. Indeed, as scientists are humans and part of the current society, they are prone to the same trappings of wielding power and influence in order to get ahead. As you get toward the end of this book, you will see me calling out increasing parts of the scientific project that are currently riddled with bias, thereby weakening the entire process and disenfranchising the most vulnerable. Reforming science is no easy task. Reforms are needed throughout the structures that fund, employ and disseminate scientific research. Sweeping this all away will take a concerted effort but might well be easier in science than in other sectors of society. Nearly all scientists recognise that science needs to be objective and impartial in order to succeed. Their objectivity makes them an ideal group to kickstart a fairer, more transparent and open system. As you will see, there has already been a move to make science more open, and by embracing this transparency the next generation of scientists will reap the benefits, and be on a sounder footing to improve a better scientific society.

@@

In his book, Demon-Haunted World, Carl Sagan [-@sagan2011demon] provides his take on how science provides scientists with what he terms ‘Baloney Detection’, commonly referred to today as ‘bullshit’. These are well worth recording here as they provide a guiding light for all scientists whether they be novices, emerging or established. They are also easily taught and the foundation of what we should spread in the Scientific Project.

As humans, we are all susceptible to bullshit, hoaxes, charlatans and contrived deceptions as they are often more appealing than the truth. However, science is about truth-seeking, and by applying a set of simple rules, Sagan argues, we can avoid being misled or distracted by ideas that may waste our time and energy by using ‘skeptical thinking’. These rules are useful in your scientific life when considering hypotheses or when building arguments to explain your results. However, as Sagan argues, these simple rules are also useful in your day to day life, and as a simplified example of the scientific method, we can pass them onto all our fellow citizens.

1. You should always seek independent confirmation of results. 
1. Seek out different opinions on the results and consider all possibilities.
1. Don’t believe facts simply because they are spoken by an authority on the subject. We know that humans are prone to [confirmation bias](#TypeI), and often exhibit motivated reasoning, especially if they have some vested interest. Even the most experienced have made mistakes in the past and will again in the future. My preferred way of thinking of this rule is to acknowledge that experts are made of ex- (a has been) and -spurt (a drip under pressure).
1. [Use a hypothetical deductive framework](howhypothesis.html). Instead of accepting the first (hypothesis) explanation, think of all the possible reasons why a result might have occurred and then consider how you could (or perhaps you already have in your experiment) disproved each one. You may find that an untested hypothesis to explain the results represents a major caveat that could be tested, or can’t yet be tested. Either way, you need to remain open to the possibility that this explains your results. This leads neatly onto:
1. Don’t get too involved with keeping your chosen hypothesis [opening you to potential cofirmation bias](#TypeI). Remain critical to the possibility that it might be wrong, and provide ideas on how to substantiate or disprove competing hypotheses. If you don’t, others will. 
1. If at all possible, use a quantitative method to test your hypothesis. You should already be well on the way to this one inside your PhD, but it is worth bearing in mind especially when it comes to considering testing other competing hypotheses. How could they be quantitatively tested?
1. Arguments ([as we see later](#argument)) consist of a chain of statements that follow one from the other. This chain is only as strong as its weakest link, and the strength of each link in the chain (your argumentative statements) must be critically considered by you. 
1. [Occam’s Razor](#argument) dictates that given two hypotheses that equally explain the results, the simpler one is likely more plausible. Or put another way, the more links you have in your chain, the more likely that one of them is not as strong as you would like it to be.
1. Your [hypothesis](#hypothesis) (or argument) must always be falsifiable. That is, that your test should be able to find that it is incorrect. If your test cannot show that your hypothesis is incorrect, then it isn’t worth much.

Sagan provides plenty of real-world examples of each of these arguments, and I’d suggest that you consult his book [@sagan2011demon] if you are interested in following these up. Similar approaches, but written in different styles, are provided by Richard Dawkins in his [-@dawkins2004devil] book: A Devil’s Chaplain, and Julia Galef in her [-@galef2021scout] book: The Scout Mindset.

@@

I would encourage you not to stop your objective and evidence-based thinking when you leave the office or laboratory. By applying the rules of science to other sectors of society, you will find that there are many societal norms that could benefit from similar reform. Certainly, a lot of the technology developed through scientific inquiry can be used to help society reform, but scientific thinking will help everyone. As well as bringing the world to scientific thinking, as scientists we should take scientific thinking to the world.

## Citizen science
For a long time, I struggled with the term 'citizen science' as I reasoned that many such ventures did not involve citizens doing science. To me, most seemed focussed on outsourcing large tasks to willing public participants: i.e. having citizens do some grunt work. Collecting data or doing a fraction of the analysis is not science. However, in light of the scientific project, I have changed my mind. All of these projects offer citizens a way to participate in scientific projects in a way that was previously unavailable to them. They promote direct communication between scientists and citizens. In this way, they serve to promote the understanding of science to the general public.

### 2021-07-20 1.04-healthy.Rmd
# Updating sections to chapters
Add files via upload
# After copy edits

## A healthy body to go with your PhD mind {#healthy}

While I’m busy handing out advice, here’s some of the best that I have to offer. If you don’t do it already, then when you start your PhD start regular exercise. There’s an old adage that ‘a healthy body leads to a healthy mind’. I never gave it much attention until I started exercising regularly some 7 years ago. Getting fit was the best thing I ever did to improve my work.

### Being physically active improves concentration

The first thing I found was that I could sit and concentrate on work for longer. Before I got fit, I was constantly distracted and needed to take lots of breaks. Now I find that my concentration is much better and that as my fitness stamina increases, so too does my ability to concentrate.

### Time to think

I used to think that the time I spent exercising was unproductive with respect to work. Now I have begun to realise that this represents some of my best thinking time. I use the time I spend trail running to turn ideas over in my head and especially to think through the logic in arguments, and potential flaws in experimental design. This isn’t to say that I don’t need to put time in at the computer writing it all down, or talking through ideas with colleagues. But I find the meditative time I spend exercising to be especially productive.

### Balancing work with life

Having already told you that doing a PhD might unbalance you mentally, regular exercise is a really good way to make sure that you retain that balance with life [see @hotaling2018publishing]. By life, I'm referring to everything away from your academic work. This might include, family, friends, sport, hobbies (even fishing). If you don't have friends outside your academic life, then it would be a good idea to find some so that you can keep in contact with the 'real world'.

If you already exercise, then you likely know how important this is. But if you don’t then my best advice (above anything else that you might read in this book) is to start as soon as you can. Do whatever is right for you. If you don’t already know, then try out different clubs or activities at your institution. There are normally lots to choose from. Make sure that you schedule regular time for your activity, even if that means that you run in the mornings before field work starts, or you spend your evenings doing tai-chi on the beach after a day at sea doing research on dolphins. Take your exercise with you wherever you go for your studies. Take it as seriously as you do your studies and you’ll find that both will benefit.

>>>>

# Keeping track of your mental health {#healthy}

Stress is a natural part of life and many people are at their most productive when they are under some degree of pressure, such as a deadline. Although deadlines don’t work for everyone.

>Douglas Adams famously claimed: 
>
>“I love deadlines. I like the whooshing sound they make as they fly by.”

Problems arise when we become overwhelmed by stress and are unable to fully respond. When this occurs, productivity can drop off and survival responses can be triggered as if responding to an actual physical attack. These responses include fight, flight or freeze responses. Anxiety and panic can be triggered. In this state, additional demands on your time may also push your life off balance, so that you start to neglect your personal wellbeing which can negatively impact on relationships, exercise regime, or even nutrition and personal hygiene. Some people can find that the additional stress can cause physical symptoms that may even need medical treatment. Your sense of competence and mastery can be negatively impacted such that you may even suffer from feelings of inadequacy or imposter syndrome ([see part 2](#fear)).

Although there are not many studies on mental health for PhD students, those that exist (as well as surveys: [Nature 2019](https://www.nature.com/articles/d41586-019-03459-7)) all suggest that there is a significant toll, which is proportionately higher than for others in society [@levecque2017work]. Whatever your prior experience of stress in a working environment, academia is known to be particularly stressful, and as a PhD student, you are likely to absorb a significant amount of this stress into your own life [@stubb2011balancing].

The General Health Questionnaire (see GHQ-12 in Table \@ref(tab:GHQ)) is an instrument used to measure psychological distress. It is quick, reliable and simple to score, so you can use it at any time during your PhD studies as an indicator of whether you need to reach out to personal, occupational or professional support networks.

Right now, I suggest you complete the GHQ-12 (Table \@ref(tab:GHQ)) and record your answers as a baseline. Keep the scores somewhere safe. During the course of your PhD, if you feel that your scores may have changed, take the test again and compare them with your baseline scores. Although there are no hard rules, if three or more of your scores have moved by two or more points it could be worth discussing with your support network to help you decide whether or not to seek professional help.

Table: (\#tab:GHQ) **A General Health Questionnaire with 12 questions (GHQ-12) that you can use to keep track of your mental health.** Answer the questions in this table now, and compare your answers to answers in your future when you are feeling more under strain. Write down any explanations you have for your choice of answers.

| General Health Questionnaire: Have you recently... | 0 | 1 | 2 | 3 |
|-|-|-|-|-|
| been feeling reasonably happy, all things considered? | Better than usual | Same as usual | Less than usual | Much less than usual |
| lost much sleep over worry? | Not at all | No more than usual | More than usual | Much more than usual | 
| been feeling unhappy and depressed? | Not at all | No more than usual | More than usual | Much more than usual |
| felt you couldn’t overcome your difficulties? | Not at all | No more than usual | More than usual | Much more than usual |
| felt under constant strain? | Not at all | No more than usual | More than usual | Much more than usual | 
| felt capable of making decisions about things? | Better than usual | Same as usual | Less than usual | Much less than usual |
| been able to face up to your problems? | Better than usual | Same as usual | Less than usual | Much less than usual | 
| been thinking of yourself as a worthless person? | Not at all | No more than usual | More than usual | Much more than usual |
| been losing confidence in yourself? | Not at all | No more than usual | More than usual | Much more than usual | 
| been able to enjoy your normal day-to-day activities? | Better than usual | Same as usual | Less than usual | Much less than usual |
| been able to concentrate on whatever you are doing? | Better than usual | Same as usual | Less than usual | Much less than usual |
| felt that you are playing a useful part in things? | Better than usual | Same as usual | Less than usual | Much less than usual |

Even if you don't feel you need the support of your institution now, it is worth finding out how they can support your mental health in the future if needed. Although there has been some stigma attached to difficulties with mental health in the past, most institutions accept that pressures are mounting on postgraduate students and that they may require support. Most institutions have experienced councillors available to support you if needed. Importantly, you should realise that none of these symptoms are unusual and that there is a high probability that many of your colleagues may also be struggling. Knowing that your problems are shared and reaching out to support networks early is an excellent way to prevent them from escalating beyond your control.

A study into the mental health of PhD students in Belgium exemplifies the kinds of difficulties that they face when compared with other similar groups (@levecque2017work; Figure \@ref(fig:Mental-health)).

(ref:Mental-health) **How does the mental health of people in different sectors compare?** A comparison of the mental health of PhD students [data from @levecque2017work] with highly educated general population, highly educated employees and higher education students using the General Health Questionnaire. The Risk Ratio (RR: adjusted for age and gender) in PhD students in Flanders, Belgium, is consistently higher (>1) when compared to any of the other surveyed groups.

```{r Mental-health, echo=FALSE, out.width='80%', fig.cap="(ref:Mental-health)"}
 knitr::include_graphics("figures/Mental-health.png")
```

No matter how well you think of your own abilities to cope with mental health issues, doing a PhD will cause you additional stress and can trigger maladaptive coping mechanisms. Learning how to cope with additional stress early in your career can be beneficial for future personal development.

Academia is recognised as a particularly stressful environment; you will likely take on some of this environmental stress in addition to any stress associated with your studies. Additional stressors come from home and family situations. Your best means of coping will be to try and develop a support network and to understand where and with whom you can discuss any difficulties as they arise. Knowing who this is and how and when to approach them will put you in a stronger position if you need them in future.

## Being physically active improves mental wellbeing

The positive relationship between the amount of physical activity and higher mental wellbeing is well established [e.g. @grasdalsmoen2020physical; @gerber2014increased], but the kind of exercise required to achieve this improved result is varied. For example, a study of college students showed that an hour of Thai Chi twice a week for 3 months was enough to have significantly improved physical and mental health scores [@wang2004effects]. Increased levels of physical activity were also found to be associated with improved sleep quality in another study [@ghrouz2019physical; @gerber2014increased]. There are plenty of studies out there that suggest there are multiple benefits from physical activity.

If you don’t do it already, then when you start your PhD start regular exercise. Getting fit was the best thing I ever did to improve my work. The first thing I found was that I could sit and concentrate on work for longer. Before I got fit, I was constantly distracted and needed to take lots of breaks. Now I find that my concentration is much better and that as my fitness stamina increases, so too does my ability to concentrate.

There are other ways to achieve these benefits, including mindfulness or meditation.

## Time to think

I used to think that the time I spent exercising was unproductive with respect to work. Now I have begun to realise that this represents some of my best thinking time. I use the time I spend trail running to turn ideas over in my head and especially to think through the logic in arguments, and potential flaws in experimental design. This isn’t to say that I don’t need to put time in at the computer writing it all down, or talking through ideas with colleagues. But I find the meditative time I spend exercising to be especially productive. Don't take my word for it, see this great new book by Caroline Williams [-@williams2021move].

## Balancing work with life

Having already told you that doing a PhD might unbalance you mentally, regular exercise is a really good way to make sure that you retain that balance with life [see @hotaling2018publishing]. By life, I'm referring to everything away from your academic work. This might include family, friends, sport, hobbies (even fishing). If you don't have friends outside your academic life, then it would be a good idea to find some so that you can keep in contact with the 'real world'.

If you already exercise, then you likely know how important this is. But if you don’t then, my best advice (above anything else that you might read in this book) is to start as soon as you can. Do whatever is right for you. If you don’t already know, then try out different clubs or activities at your institution. There are normally lots to choose from. Make sure that you schedule regular time for your activity, even if that means that you run in the mornings before field work starts, or you spend your evenings doing tai-chi on the beach after a day at sea doing research on dolphins. Take your exercise with you wherever you go for your studies. Take it as seriously as you do your studies and you’ll find that both will benefit.
